The illusory promise of the Aligned Rank Transform

### 2025-05-25 sandbox.qmd
corrections in intro and background

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "Appendix I. Additional experimental results"
author: 
  - name: Theophanis Tsandilas
    email: theophanis.tsandilas@inria.fr
    orcid: 0000-0002-0158-228X
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiez@univ-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib

@@

Used for quick testing... eeee

>>>>

---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "A systematic study of rank transformations"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib
csl: canadian-journal-of-philosophy.csl

@@

::: {.callout-important appearance="simple"}
## Under Review {.unnumbered}
This paper is [under review](https://www.journalovi.org/under-review.html) on the experimental track of
the [Journal of Visualization and Interaction](https://www.journalovi.org/). See the 
[reviewing process](https://github.com/journalovi/2024-tsandilas-ranktransforms/issues).
:::

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Useful libraries
library(crosstalk)
library(kableExtra)
library(gridExtra)
library(lmerTest)
library(tidyverse)
library(plotly)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Source code for reading data from experimental results and for plotting
source("dataReaders.R")
source("plotting.R")
source("effects_plot.R")
source("interactions_plot.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

::: {.callout-note appearance="simple" icon=false collapse=false}

{{< include text/10-abstract.qmd >}}

::: {#tbl-example}
```{r, echo=FALSE, warning=FALSE}
df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
```

::: {#fig-example}
```{r, echo=FALSE, warning=FALSE, fig.height=2.7, fig.width=8}
cbPalette <- c("#999999", "#E69F00", "#F15854")

p1 <- (df %>%  group_by(Participant,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1, 0)) +
        theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))

p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 6) +
        #geom_jitter(shape=20, position=position_jitter(0.1)) +
        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
        theme_bw() + theme(legend.position = "none") +
        #scale_fill_brewer()
        scale_fill_manual(values=cbPalette))

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```
Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
:::

::: {#fig-example-ordinal}
```{r, echo=FALSE, warning=FALSE, fig.height=3, fig.width=9}
df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))

p3 <- (df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Perceived Efficiency") +
        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
        theme(axis.text.y=element_text(angle=50, hjust=0.7))

#diffPalette <- c("#f9f9f9", "#e2e2e2", "#b9b9b9", "#999999")
diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
p4 <- (df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank())

grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
```
Boxplots summarizing the results on efficiency as perceived by participants. Circles represent the mean rating of each individual participant. 
:::

### 2025-05-25 supplementary/examples-case-studies/1-Introduction/illustrative-example.Rmd
corrections in intro and background

The experiment follows a fully balanced repeated-measures factorial design, where each participant ($N = 12$) performs 12 tasks in a unique order. The researcher measures the time that it takes participants to complete each task.

@@

df$Participant <- factor(df$Participant)
df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
df$Technique <- factor(df$Technique)
```

@@

p1 <- (df %>%  group_by(Participant,Technique) %>% 
summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette)

@@

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```

### Statistical analysis

We conduct an analysis with five different methods: PAR, LOG, RNK, ART, and INT.
```{r, warning=FALSE}
# Implementation of the Inverse Normal Transform
int = function(x){
   qnorm((rank(x) - 0.5)/length(x))
}

@@

### Main and interaction effects
We finally report the results of our analysis:
```{r, warning=FALSE}
# Function for printing the results for each method
printAnalysis <- function(num, method, model) {
  cat(num, ". Analysis using ", method, "\n\n", sep="")

@@

### Contrasts
Finally, we conduct an analysis of pairwise differences between the three techniques *A*, *B*, and *C*, using a Bonferroni correction. For ART, we use the ```art.con``` function.

@@

>>>>

The experiment follows a fully balanced repeated-measures factorial design, where each participant ($N = 12$) performs 12 tasks in a unique order. The researchers measure the time that it takes participants to complete each task, but participants also evaluate the perceived efficiency, using an ordinal scale with five levels.

@@

df$Participant <- factor(df$Participant)
df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
df$Technique <- factor(df$Technique)

df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))
```

@@

# Plots for the time measure 
p1 <- (df %>%  group_by(Participant,Technique) %>% 
summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette)

@@

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))

# Plots for the perceived efficiency measure
dd1 <- df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop")

p3 <- (dd1 %>%   
        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Perceived Efficiency (means)") +
        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
        theme(axis.text.y=element_text(angle=50, hjust=0.7))

dd2 <- df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop")
diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
p4 <- (dd2 %>%   
        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank())

grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
```

### Statistical analysis of time performance

For task-completion times, we conduct an analysis with five different methods: PAR, LOG, RNK, ART, and INT.
```{r, warning=FALSE}
# Implementation of the Inverse Normal Transform
int = function(x){
   qnorm((rank(x) - 0.5)/length(x))
}

@@

#### Main and interaction effects
We report the results of our analysis.
```{r, warning=FALSE}
# Function for printing the results for each method
printAnalysis <- function(num, method, model) {
  cat(num, ". Analysis using ", method, "\n\n", sep="")

@@

We observe that only ART shows statistical evidence for a main effect of Technique, as well as an interaction effect.

#### Contrasts
Finally, we conduct an analysis of pairwise differences between the three techniques *A*, *B*, and *C*, using a Bonferroni correction. For ART, we use the ```art.con``` function.

@@

### Statistical analysis of perceived efficiency
We conduct an analysis with PAR, RNK, ART, and INT.
```{r, warning=FALSE}
df$Perceived_Efficiency <- as.integer(df$Perceived_Efficiency)

mpar <- lmer(Perceived_Efficiency ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
mart <- art(Perceived_Efficiency ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
mrnk <- lmer(rank(Perceived_Efficiency) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
mint <- lmer(int(Perceived_Efficiency) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
```

We report the results of our analysis.
```{r, warning=FALSE}
printAnalysis(2, "ART", mart)
printAnalysis(3, "Rank Transformation", mrnk)
printAnalysis(4, "Inverse Normal Transformation", mint)
printAnalysis(5, "No Transformation", mpar)
```

ART is the only method to detect an main effect for both Difficulty ($p=.0011$, partial $\eta^2 = .12$, $95\%$ CI $[.03, 1.0]$) and Technique ($p=.00016$, partial $\eta^2 = .13$, $95\%$ CI $[.05, 1.0]$). Its *p-*values are substantially lower than the *p-*values of the other methods, and its effect size estimates are surprisingly large.

### 2025-05-25 text/20-intro.qmd
corrections in intro and background

Suppose an HCI researcher conducts an experiment to compare the performance of three user interface techniques (A, B, and C) that help users complete image editing tasks of four different difficulty levels. The experiment follows a fully balanced $4 \times 3$ repeated-measures factorial design, where each participant (N = 12) performs 12 tasks in a unique order. The researcher measures the time that it takes participants to complete each task. The following table presents the experimental results:

::: {#tbl-example}
```{r, echo=FALSE, warning=FALSE}
df <- read.csv("example_data.csv", sep=",", header=TRUE, strip.white=TRUE)
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
``` 
Example dataset: Time (in minutes) spent by 12 participants for four difficulty levels and three user interface techniques. Scroll down to see the full results.
:::

The experiment is hypothetical but has similarities with real-world experiments, e.g., see the experiments of @Fruchard:2023. Time performances have been randomly sampled from a population in which: (1) *Difficulty* has a large effect; (2) *Technique* has no effect; and (3) there is no interaction effect between the two factors. To generate time values, we drew samples from a log-normal distribution. The log-normal distribution is often a good fit for real-world measurements that are bounded by zero and have low means but large variance [@Limbert:2001]. Task-completion times are good examples of such measurements [@Sauro:2010].

@@

p1 <- (df %>%  group_by(Participant,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))

@@

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```
Boxplots summarizing the results of our illustrative example. Dots represent the median time performance of each individual participant. 
:::

@@

We observe that ART exaggerates both the effect of *Technique* and its interaction with *Difficulty*.

### Overview
The above example does not capture a rare phenomenon. We will show that ART's error inflation is systematic for a range of distributions that deviate from normality, both continuous and discrete. We will explain how the problem emerges. We will also see that ART often performs worse than simpler methods that ART is widely considered to repair or improve, such as the pure rank transformation or no transformation at all.

>>>>

Suppose a team of HCI researchers conduct an experiment to compare the performance of three user interface techniques (A, B, and C) that help users complete image editing tasks of four different difficulty levels. The experiment follows a fully balanced $4 \times 3$ repeated-measures factorial design, where each participant (N = 12) performs 12 tasks in a unique order. The researchers measure the time that it takes participants to complete each task. In addition, the participants rate their perceived efficiency completing each task, using an ordinal scale with five levels: (1) *inefficient*, (2) *somewhat inefficient*, (3) *neutral*, (4) *somewhat efficient*, and (5) *efficient*.

The following table presents the experimental results:

::: {#tbl-example}
```{r, echo=FALSE, warning=FALSE}
df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
``` 
Example dataset: Time (in minutes) spent by 12 participants and their perceived efficiency (ordinal scale) for four difficulty levels and three user interface techniques. Scroll down for full results.
:::

The experiment is hypothetical but has similarities with real-world experiments, e.g., see the experiments of @Fruchard:2023.

**Time performance.** Time measurements have been randomly sampled from a population in which: (1) *Difficulty* has a large main effect; (2) *Technique* has no main effect; and (3) there is no interaction effect between the two factors. To generate time values, we drew samples from a log-normal distribution. The log-normal distribution is often a good fit for real-world measurements that are bounded by zero and have low means but large variance [@Limbert:2001]. Task-completion times are good examples of such measurements [@Sauro:2010].

@@

p1 <- (df %>%  group_by(Participant,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1, 0)) +
        theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))

@@

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```
Boxplots summarizing time performance results for our illustrative example. Dots represent the median time for each individual participant. 
:::

@@

We observe that ART exaggerates both the effect of *Technique* and its interaction with *Difficulty*. This example demonstrates a more general problem with ART’s alignment mechanism under long-tailed distributions. ART tends to confound effects: a strong effect on a single factor often leads the method to detect spurious effects on other factors or to identify non-existent interactions.

**Perceived efficiency.**  The perceived efficiency ratings were randomly sampled from a population with no main effect on any of the two factors and no interaction effect. @fig-example-ordinal shows that participants' ratings are concentrated at the higher end of the scale but with no clear trends.

::: {#fig-example-ordinal}
```{r, echo=FALSE, warning=FALSE, fig.height=2.7, fig.width=8}
df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))

p3 <- (df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Perceived Efficiency") +
        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
        theme(axis.text.y=element_text(angle=50, hjust=0.7))

#diffPalette <- c("#f9f9f9", "#e2e2e2", "#b9b9b9", "#999999")
diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
p4 <- (df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank())

grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
```
Boxplots summarizing the results on efficiency as perceived by participants. Circles represent the mean rating of each individual participant. 
:::

For our analysis, we compare the results of PAR, ART, RNK, and INT, since LOG is not relevant for this type of data. We also report the results of the Friedman test [@Friedman] for each factor, where we first aggregate the data by taking the mean over the other factor. Results are as follows:

|        | PAR  | ART | RNK | INT | Friedman Test |
|--------|------|-----|-----|-----|-----|
| Difficulty  | $.93$ | $.0011$  |  $.86$ | $.88$ | $.83$ |
| Technique   | $.18$ | $.00016$  | $.32$ | $.23$ | $.24$ |
| Difficulty $\times$ Technique | $.27$ | $.10$ | $.33$ | $.34$ | |
: *p*-values for main and interaction effects {.sm}

ART's *p*-values for the two main effects are surprisingly low, while no other methods shows any substantial evidence for such effects. We also presents the methods' effect size estimates:

|        | PAR  | ART | RNK | INT |
|--------|------|-----|-----|-----|
| Difficulty  | $.004\ [.00, 1.0]$  | $.12\ [.03, 1.0]$  |  $.006\ [.00, 1.0]$ | $.005\ [.00, 1.0]$ |
| Technique   | $.03\ [.00, 1.0]$ | $.13\ [.05, 1.0]$  | $.02\ [.00, 1.0]$ | $.02\ [.00, 1.0]$ |
| Difficulty $\times$ Technique | $.06\ [.00, 1.0]$ | $.04\ [.00, 1.0]$  | $.05\ [.00, 1.0]$| $.05\ [.00, 1.0]$|
: partial $\eta^2$ and its 95\% confidence interval {.sm}

The discrepancies between ART's estimates for the two main effects and those of the other methods are large. As we demonstrate in this paper, such discrepancies stem from ART's inability to handle discrete distributions. The issue becomes more pronounced with larger sample sizes. @luepsen:2017 previously warned researchers about this problem, but his warnings have largely been ignored.

### Overview
The above example does not reflect a rare phenomenon. We will show that ART's error inflation is systematic for a range of distributions that deviate from normality, both continuous and discrete.

### 2025-05-25 text/30-background.qmd
corrections in intro and background

## Background {#background}
In this section, we provide background on the history, scope, construction, and use of rank transformation methods, with a particular focus on ART. Additionally, we summarize the results of previous evaluation studies and position our own work within this context.

@@

### Rank transformations
Unfortunately, the scope of nonparametric tests is limited, as they only support simple experimental designs with a single independent variable. Rank transformations aim to address this limitation by bridging the gap between nonparametric statistics and ANOVA.

@@

**The aligned rank transform.** Given such negative results, many researchers turned to aligned (or adjusted) rank transformations. @Sawilowsky:1990 discusses several variations of aligned rank-based transformations (ART) and tests, while @higgins:1990 detail the ART method that we evaluate in this paper for two-way designs. @wobbrock:2011 generalize it to more than two factors, while 10 years later, @elkin:2021 show how to apply it to multifactor contrast tests. @fig-ART-explained explains the general method for a design with two factors, $A$ and $B$.

@@

The key intuition behind the transform is that responses are ranked after *aligning* them, such that effects of interest are separated from other effects. This means that for each effect of interest, either main or interaction, we produce a separate ranking. This is clearly demonstrated through the example dataset in @fig-ART-explained (a), where each aligned ranking ($ART_A$, $ART_B$, and $ART_{AB}$) is for testing a different effect ($A$, $B$, and $A \times B$, respectively).

Notice that even for this very small dataset ($n = 2$), these three rankings are distinct from each other, as well as very different from the ranking of the simple rank transform. Interestingly, equal responses (the two $0.3$ in our example dataset) may be assigned to very different ranks, while very different responses (such as $0.3$ and $3.0$ in our dataset) can be assigned to the same rank. Furthermore, differences between the two groups $a_1$ and $a_2$ within factor $A$ are significantly exaggerated by ART (see $ART_A$ ranks). We will later see that this behavior can lead to unstable results and conclusions, and is the source of the main problems of the method. For example, if we run a repeated-measures ANOVA on these ranks, we obtain a *p*-value $= .044$ for the effect of $A$. For comparison, if we run repeated-measures ANOVA on the ranks of simple rank transformation, we obtain a *p*-value $= .46$, while if we use a log-transformation that is more appropriate for these data, we obtain a *p*-value $= .85$.

@@

This is not the only alignment technique found in the research literature. @Sawilowsky:1990 suggests that at least for balanced designs, interactions could be removed when aligning main effects, in the same way main effects are removed when aligning interactions. This approach is also taken by @Leys:2010, who derived a common ranking for both main effects after subtracting the interaction term. We do not test these alternative alignment methods in this paper since to the best of our knowledge, they are not commonly used.

**The inverse normal transform.** 
A third transformation method that we evaluate is the rank-based inverse normal transformation (INT). INT is more than 70 years old [@VanDerWaerden:1952] and appears in several variations [@Beasley:2009;@Solomon2009]. Its general formulation is as follows:

@@

**Other non-parametric rank-based methods.** There are other rank-based statistical tests that deal with interactions, the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] being the most representative one. @Kaptein:2010 introduced this method to the HCI community, advocating its application in the analysis of Likert-type data as a viable alternative to parametric ANOVA. In addition to ATS, Lüpsen [-@luepsen:2017; -@luepsen:2018; -@luepsen:2023] investigated several other multifactorial nonparametric methods. In particular, the author evaluated the hybrid ART+INT technique proposed by @Mansouri:1995, which applies INT on the ranks of ART. He also tested multifactorial generalizations of the van der Waerden test [@VanDerWaerden:1952] and the Kruskal-Wallis and Friedman tests [@Kruskal-Wallis;@Friedman]. The former is based on INT, but instead of using F-tests on the transformed values as part of ANOVA, it computes $\chi^2$ ratios over sums of squares. These two methods are not widely available. Their implementation in R can be downloaded from Lüpsen's web page [@luepsen_R].

### Experimental evaluations
Previous studies have assessed ART and related procedures through various Monte Carlo experiments, yielding conflicting results and conclusions.

**Results in support of ART**. The outcomes from numerous experiments conducted in the 80s and 90s suggested that ART was robust in testing interaction effects. Noteworthy instances include studies, such as those by @Salter:1993, which compared the method to parametric ANOVA. The authors found that ART remains robust even in the presence of outliers or specific non-normal distributions, such as the logistic, exponential, and double exponential distributions. Their findings indicated only a marginal increase in error rates (ranging from 6.0% to 6.3% instead of the expected 5%) when applied to the exponential distribution. Furthermore, ART demonstrated superior statistical power compared to parametric ANOVA. @Mansouri:1995 evaluated ART under a different set of non-normal distributions (normal, uniform, log-normal, exponential, double exponential, and Cauchy) in the presence of increasing main effects. Except for the Cauchy distribution, ART maintained Type I error rates close to nominal levels across all scenarios, irrespective of the magnitude of main effects. In contrast, the error rates of the rank transformation reached very high levels (up to $100\%$) as the magnitude of main effects increased, even under the normal distribution. ART only failed under the Cauchy distribution, which is well-known to be pathological.

>>>>

## Background {#background}
We provide background on the history, scope, construction, and use of rank transformation methods, with a particular focus on ART. Additionally, we summarize the results of previous evaluation studies and position our own work within this context.

@@

### Rank transformations
 The scope of nonparametric tests is limited, as they only support simple experimental designs with a single independent variable. Rank transformations aim to address this limitation by bridging the gap between nonparametric statistics and ANOVA.

@@

**The aligned rank transform.** In response to these negative results, many researchers turned to aligned (or adjusted) rank transformations. @Sawilowsky:1990 discusses several variations of aligned rank-based transformations (ART) and tests, while @higgins:1990 describe the specific ART method that we evaluate in this paper for two-way designs. @wobbrock:2011 generalize it to more than two factors, and a decade later, @elkin:2021 show how to apply it to multifactor contrast tests. @fig-ART-explained explains the general method for a design with two factors, $A$ and $B$.

@@

The key intuition behind the transform is that responses are ranked after they are *aligned*, such that effects of interest are separated from other effects. This implies that for each effect, either main or interaction, a separate ranking is produced. This is clearly illustrated in the example dataset shown in @fig-ART-explained (a), where each aligned ranking ($ART_A$, $ART_B$, and $ART_{AB}$) is for testing a different effect ($A$, $B$, and $A \times B$, respectively).

Notice that even for this very small dataset ($n = 2$), the three ART rankings are distinct from one another, and also markedly different from the ranking produced by a simple rank transformation. Interestingly, identical responses (e.g., the two $0.3$ in our example dataset) may be assigned very different ranks, while responses that differ greatly (e.g., $0.3$ and $3.0$) can be assigned the same rank. Furthermore, ART significantly exaggerates the differences between the two groups $a_1$ and $a_2$ within factor $A$ (as seen in the $ART_A$ ranks).

As we will show later, this behavior can lead to unstable results and misleading conclusions, and it represents a central flaw of the method. For example, running a repeated-measures ANOVA on these ART ranks yields a *p*-value of $.044$ for the effect of $A$. In contrast, using a repeated-measures ANOVA on the ranks of simple rank transformation produces a *p*-value of $.46$, while applying a log-transformation --- more appropriate for this type of data --- yields a *p*-value of $.85$.

@@

This is not the only alignment technique discussed in the literature. @Sawilowsky:1990 suggests that, at least for balanced designs, interactions could also be removed when aligning main effects, in the same way main effects are removed when aligning interactions. This approach is also taken by @Leys:2010, who derived a common ranking for both main effects after subtracting the interaction term. We do not evaluate these alternative alignment methods in this paper, as they are, to the best of our knowledge, not commonly used in practice.

**The inverse normal transform.** 
A third transformation method we evaluate is the rank-based inverse normal transformation (INT). INT has been in use for over 70 years [@VanDerWaerden:1952] and exists in several variations [@Beasley:2009;@Solomon2009]. Its general formulation is as follows:

@@

**Other non-parametric rank-based methods.** Several other rank-based statistical tests handle interactions, with the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] being the most representative one. @Kaptein:2010 introduced this method to the HCI community, advocating its use for analyzing Likert-type data as a viable alternative to parametric ANOVA. In addition to ATS, Lüpsen [-@luepsen:2017; -@luepsen:2018; -@luepsen:2023] investigated several other multifactorial nonparametric methods. In particular, the author evaluated the hybrid ART+INT technique proposed by @Mansouri:1995, which applies INT on the ranks of ART. He also tested multifactorial generalizations of the van der Waerden test [@VanDerWaerden:1952] and the Kruskal-Wallis and Friedman tests [@Kruskal-Wallis;@Friedman]. The former is based on INT, but instead of using F-tests on the transformed values as part of ANOVA, it computes $\chi^2$ ratios over sums of squares. These two methods are not widely available, but implementations in R can be downloaded from Lüpsen's website [@luepsen_R].

### Experimental evaluations
Previous studies have evaluated ART and related procedures using various Monte Carlo simulations, often producing conflicting results and conclusions.

**Results in support of ART**. A number of experiments conducted during the 80s and 90s suggested that ART is robust for testing interaction effects. Noteworthy instances include studies, such as those by @Salter:1993, which compared the method to parametric ANOVA. The authors found that ART remains robust even in the presence of outliers or specific non-normal distributions, such as the logistic, exponential, and double exponential distributions. Their findings indicated only a marginal increase in error rates (ranging from 6.0% to 6.3% instead of the expected 5%) when applied to the exponential distribution. Furthermore, ART demonstrated superior statistical power compared to parametric ANOVA. @Mansouri:1995 evaluated ART under a different set of non-normal distributions (normal, uniform, log-normal, exponential, double exponential, and Cauchy) in the presence of increasing main effects. Except for the Cauchy distribution, ART maintained Type I error rates close to nominal levels across all scenarios, irrespective of the magnitude of main effects. In contrast, the error rates of the rank transformation reached very high levels (up to $100\%$) as the magnitude of main effects increased, even under the normal distribution. ART only failed under the Cauchy distribution, which is well-known to be pathological.
